---
title: "Adding Custom Feature Engineering Functions"
author: "Jenna Reps, Egill Fridgeirsson"
date: "`r Sys.Date()`"
header-includes:
    - \usepackage{fancyhdr}
    - \pagestyle{fancy}
    - \fancyhead{}
    - \fancyhead[CO,CE]{Custom Patient-Level Prediction Algorithms}
    - \fancyfoot[CO,CE]{PatientLevelPrediction Package Version `r    utils::packageVersion("PatientLevelPrediction")`}
    - \fancyfoot[LE,RO]{\thepage}
    - \renewcommand{\headrulewidth}{0.4pt}
    - \renewcommand{\footrulewidth}{0.4pt}
output:
  pdf_document:
    number_sections: yes
    toc: yes
  html_document:
    number_sections: yes
    toc: yes
---

```{=html}
<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Adding Custom Feature Engineering}
-->
```
```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(PatientLevelPrediction)
```

# Introduction

This vignette describes how you can add your own custom function for feature engineering in the Observational Health Data Sciences and Informatics (OHDSI) [`PatientLevelPrediction`](http://github.com/OHDSI/PatientLevelPrediction) package. This vignette assumes you have read and are comfortable with building single patient level prediction models as described in the [`BuildingPredictiveModels` vignette](https://github.com/OHDSI/PatientLevelPrediction/blob/main/inst/doc/BuildingPredictiveModels.pdf).

**We invite you to share your new feature engineering functions with the OHDSI community through our [GitHub repository](http://github.com/OHDSI/PatientLevelPrediction).**

# Feature Engineering Function Code Structure

To make a custom feature engineering function that can be used within PatientLevelPrediction you need to write two different functions. The 'create' function and the 'implement' function.

The 'create' function, e.g., create\<FeatureEngineeringFunctionName\>, takes the parameters of the feature engineering 'implement' function as input, checks these are valid and outputs these as a list of class 'featureEngineeringSettings' with the 'fun' attribute specifying the 'implement' function to call.

The 'implement' function, e.g., implement\<FeatureEngineeringFunctionName\>, must take as input:

-   `trainData` - a list containing:

    -   `covariateData`: the `plpData$covariateData`restricted to the training patients

    -   `labels`: a data frame that contain `rowId`(patient identifier) and `outcomeCount` (the class labels)

    -   `folds`: a data.frame that contains `rowId` (patient identifier) and `index` (the cross validation fold)

-   `featureEngineeringSettings` - the output of your create\<FeatureEngineeringFunctionName\>

The 'implement' function can then do any manipulation of the `trainData` (adding new features or removing features) but must output a `trainData` object containing the new `covariateData`, `labels` and `folds` for the training data patients.

# Example

Let's consider the situation where we wish to create an age spline feature. To make this custom feature engineering function we need to write the 'create' and 'implement' R functions.

## Create function

Our age spline feature function will create a new feature using the `plpData$cohorts$ageYear` column. We will implement a restricted cubic spline that requires specifying the number of knots. Therefore, the inputs for this are: `knots` - an integer/double specifying the number of knots.

```{r, echo = TRUE, eval=FALSE}
createAgeSpline <- function(
                     knots = 5
                     ){
  
  # create list of inputs to implement function
  featureEngineeringSettings <- list(
    knots = knots
    )
  
  # specify the function that will implement the sampling
  attr(featureEngineeringSettings, "fun") <- "implementAgeSplines"

  # make sure the object returned is of class "sampleSettings"
  class(featureEngineeringSettings) <- "featureEngineeringSettings"
  return(featureEngineeringSettings)
  
}
```

We now need to create the 'implement' function `implementAgeSplines()`

## Implement function

All 'implement' functions must take as input the `trainData` and the `featureEngineeringSettings` (this is the output of the 'create' function). They must return a `trainData` object containing the new `covariateData`, `labels` and `folds`.

In our example, the `createAgeSpline()` will return a list with 'knots'. The `featureEngineeringSettings` therefore contains this.

```{r tidy=FALSE,eval=FALSE}
implementAgeSplines <- function(trainData, featureEngineeringSettings, model=NULL) {
  # if there is a model, it means this function is called through applyFeatureengineering, meaning it   # should apply the model fitten on training data to the test data
  if (is.null(model)) {
    knots <- featureEngineeringSettings$knots
    ageData <- trainData$labels
    y <- ageData$outcomeCount
    X <- ageData$ageYear
    model <- mgcv::gam(
      y ~ s(X, bs='cr', k=knots, m=2)
    )
    newData <- data.frame(
      rowId = ageData$rowId,
      covariateId = 2002,
      covariateValue = model$fitted.values
    )
  }
  else {
    ageData <- trainData$labels
    X <- trainData$labels$ageYear
    y <- ageData$outcomeCount
    newData <- data.frame(y=y, X=X)
    yHat <- predict(model, newData)
    newData <- data.frame(
      rowId = trainData$labels$rowId,
      covariateId = 2002,
      covariateValue = yHat
    )
  }
  
  # remove existing age if in covariates 
  trainData$covariateData$covariates <- trainData$covariateData$covariates |> 
    dplyr::filter(!covariateId %in% c(1002))
  
  # update covRef
  Andromeda::appendToTable(trainData$covariateData$covariateRef, 
                           data.frame(covariateId=2002,
                                      covariateName='Cubic restricted age splines',
                                      analysisId=2,
                                      conceptId=2002))
  
  # update covariates
  Andromeda::appendToTable(trainData$covariateData$covariates, newData)
  
  featureEngineering <- list(
    funct = 'implementAgeSplines',
    settings = list(
      featureEngineeringSettings = featureEngineeringSettings,
      model = model
    )
  )
  
  attr(trainData$covariateData, 'metaData')$featureEngineering = listAppend(
    attr(trainData$covariateData, 'metaData')$featureEngineering,
    featureEngineering
  )
  
  return(trainData)
}

```

# Acknowledgments

Considerable work has been dedicated to provide the `PatientLevelPrediction` package.

```{r tidy=TRUE,eval=TRUE}
citation("PatientLevelPrediction")
```

**Please reference this paper if you use the PLP Package in your work:**

[Reps JM, Schuemie MJ, Suchard MA, Ryan PB, Rijnbeek PR. Design and implementation of a standardized framework to generate and evaluate patient-level prediction models using observational healthcare data. J Am Med Inform Assoc. 2018;25(8):969-975.](http://dx.doi.org/10.1093/jamia/ocy032)

This work is supported in part through the National Science Foundation grant IIS 1251151.
