---
title: "Adding Custom Sampling Functions"
author: "Jenna Reps"
date: "`r Sys.Date()`"
header-includes:
    - \usepackage{fancyhdr}
    - \pagestyle{fancy}
    - \fancyhead{}
    - \fancyhead[CO,CE]{Custom Patient-Level Prediction Algorithms}
    - \fancyfoot[CO,CE]{PatientLevelPrediction Package Version `r    utils::packageVersion("PatientLevelPrediction")`}
    - \fancyfoot[LE,RO]{\thepage}
    - \renewcommand{\headrulewidth}{0.4pt}
    - \renewcommand{\footrulewidth}{0.4pt}
output:
  pdf_document:
    number_sections: yes
    toc: yes
  html_document:
    number_sections: yes
    toc: yes
---
<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Adding Custom Sampling}
-->

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(PatientLevelPrediction)
```
# Introduction

This vignette describes how you can add your own custom function for sampling the target population in the Observational Health Data Sciencs and Informatics (OHDSI) [`PatientLevelPrediction`](http://github.com/OHDSI/PatientLevelPrediction) package. This vignette assumes you have read and are comfortable with building single patient level prediction models as described in the  [`BuildingPredictiveModels` vignette](https://github.com/OHDSI/PatientLevelPrediction/blob/master/inst/doc/BuildingPredictiveModels.pdf).

**We invite you to share your new sample functions with the OHDSI community through our [GitHub repository](http://github.com/OHDSI/PatientLevelPrediction).**

# Sample Function Code Structure 

To make a sampling function that can be used within PatientLevelPrediction you need to write two different functions.  The 'create' function and the 'implement' function.  

The 'create' function, e.g., create\<SampleFunctionName\>, takes the parameters of the sample 'implement' function as input, checks these are valid and outputs these as a list of class 'sampleSettings' with the 'fun' attribute specifying the 'implement' function to call.

The 'implement' function, e.g., implement\<SampleFunctionName\>,  must take as input:
 * trainData - a list containing:
   - covariateData: the plpData$covariateData restricted to the training patients
   - labels: a data frame that contain rowId (patient identifier) and outcomeCount (the class labels)
   - folds: a data.frame that contains rowId (patient identifier) and index (the cross validation fold)
 * sampleSettings - the output of your create\<SampleFunctionName\>
 
The 'implement' function can then do any manipulation of the trainData (such as undersampling or oversampling) but must output a trainData object containing the covariateData, labels and folds for the new training data sample.

# Example

Let's consider the situation where we wish to take a random sample of the training data population.  To make this custom sampling function we need to write the 'create' and 'implement' R functions.

## Create function

Our random sampling function will randomly sample `n` patients from the trainData.  Therefore, the inputs for this are: 
 * `n` an integer/double specifying the number of patients to sample
 * `sampleSeed` an integer/double specifying the seed for reproducibility
 
```{r, echo = TRUE, eval=FALSE}
createRandomSampleSettings <- function(
                     n = 10000,
                     sampleSeed = sample(10000,1)
                     ){
  
  # add input checks
  checkIsClass(n, c('numeric','integer'))
  checkHigher(n,0)
  checkIsClass(sampleSeed, c('numeric','integer'))
  
  # create list of inputs to implement function
  sampleSettings <- list(
    n = n,
    sampleSeed  = sampleSeed 
    )
  
  # specify the function that will implement the sampling
  attr(sampleSettings, "fun") <- "implementRandomSampleSettings"

  # make sure the object returned is of class "sampleSettings"
  class(sampleSettings) <- "sampleSettings"
  return(sampleSettings)
  
}
```

We now need to create the 'implement' function `implementRandomSampleSettings()`

## Implement function

All 'implement' functions must take as input the trainData and the sampleSettings (this is the output of the 'create' function).  They must return a trainData object containing the covariateData, labels and folds.

In our example, the `createRandomSampleSettings()` will return a list with 'n' and 'sampleSeed'.  The sampleSettings therefore contains these.

```{r tidy=FALSE,eval=FALSE}
implementRandomSampleSettings <- function(trainData, sampleSettings){

  n <- sampleSetting$n
  sampleSeed <- sampleSetting$sampleSeed
  
  if(n > nrow(trainData$labels)){
    stop('Sample n bigger than training population')
  }
  
  # set the seed for the randomization
  set.seed(sampleSeed)
  
  # now implement the code to do your desired sampling
  
  sampleRowIds <- sample(trainData$labels$rowId, n)
  
  sampleTrainData <- list()
  
  sampleTrainData$labels <- trainData$labels %>% 
    dplyr::filter(.data$rowId %in% sampleRowIds) %>% 
    dplyr::collect()
  
  sampleTrainData$folds <- trainData$folds %>% 
    dplyr::filter(.data$rowId %in% sampleRowIds) %>% 
    dplyr::collect()
  
  sampleTrainData$covariateData <- Andromeda::andromeda()
  sampleTrainData$covariateData$covariateRef <-trainData$covariateData$covariateRef
  sampleTrainData$covariateData$covariates <- trainData$covariateData$covariates %>% dplyr::filter(.data$rowId %in% sampleRowIds)
  
  #update metaData$populationSize 
  metaData <- attr(trainData$covariateData, 'metaData')
  metaData$populationSize = n
  attr(sampleTrainData$covariateData, 'metaData') <- metaData
  
  # make the cocvariateData the correct class
  class(sampleTrainData$covariateData) <- 'CovariateData'
  
  # return the updated trainData
  return(sampleTrainData)
}

```


# Acknowledgments

Considerable work has been dedicated to provide the `PatientLevelPrediction` package.

```{r tidy=TRUE,eval=TRUE}
citation("PatientLevelPrediction")
```

**Please reference this paper if you use the PLP Package in your work:**

[Reps JM, Schuemie MJ, Suchard MA, Ryan PB, Rijnbeek PR. Design and implementation of a standardized framework to generate and evaluate patient-level prediction models using observational healthcare data. J Am Med Inform Assoc. 2018;25(8):969-975.](http://dx.doi.org/10.1093/jamia/ocy032)


This work is supported in part through the National Science Foundation grant IIS 1251151.



