---
title: "Building multiple patient-level predictive models"
author: "Jenna Reps, Martijn J. Schuemie, Patrick B. Ryan, Peter R. Rijnbeek"
date: '`r Sys.Date()`'
output:
  pdf_document:
    includes:
      in_header: preamble.tex
    number_sections: yes
    toc: yes
  word_document:
    toc: yes
  html_document:
    number_sections: yes
    toc: yes
---
  <!--
  %\VignetteEngine{knitr}
%\VignetteIndexEntry{Building multiple patient-level predictive models}
-->

\newpage
# Introduction
This vignette describes how you can use the `PatientLevelPrediction` package to build multiple patient-level predictive models. This vignette assumes you have read and are comfortable with running single patient level prediction models.


# Study Populations
The create a study population setting use the function `createStudyPopulationSettings`.  If we want to test out three study populations:

- study population 1 includes people who have the outcome but leave the database before the end of time-at-risk and only those without the outcome who are observed for the whole time-at-risk period.
- study population 2 includes people who are observed for the whole time-at-risk period.
- study population 3 includes people who are not observed for the whole time-at-risk

we can make all three populations and then combine them into a list:

```{r tidy=FALSE,eval=FALSE}
studyPop1 <- createStudyPopulationSettings(binary = T,
                                          includeAllOutcomes = T,
                                          removeSubjectsWithPriorOutcome = TRUE,
                                          priorOutcomeLookback = 99999,
                                          requireTimeAtRisk = T,
                                          minTimeAtRisk=364,
                                          riskWindowStart = 1,
                                          riskWindowEnd = 365,
                                          verbosity = "INFO")
studyPop2 <- createStudyPopulationSettings(binary = T,
                                           includeAllOutcomes = F,
                                           removeSubjectsWithPriorOutcome = TRUE,
                                           priorOutcomeLookback = 99999,
                                           requireTimeAtRisk = T,
                                           minTimeAtRisk=364,
                                           riskWindowStart = 1,
                                           riskWindowEnd = 365,
                                           verbosity = "INFO")
studyPop3 <- createStudyPopulationSettings(binary = T,
                                           includeAllOutcomes = F,
                                           removeSubjectsWithPriorOutcome = TRUE,
                                           priorOutcomeLookback = 99999,
                                           requireTimeAtRisk = F,
                                           minTimeAtRisk=364,
                                           riskWindowStart = 1,
                                           riskWindowEnd = 365,
                                           verbosity = "INFO")
                                           
populationSettingList <- list(studyPop1,studyPop2,studyPop3)
```
# Covariate Settings   
The covariate settings are created using `createCovariateSettings`.  We create seperate covariate settings and then combine them into a list:

```{r tidy=FALSE,eval=FALSE}
covSet1 <- createCovariateSettings(useDemographicsGender = T, 
                                   useDemographicsAgeGroup = T, 
                                   useConditionGroupEraAnyTimePrior = T,
                                   useDrugGroupEraAnyTimePrior = T)
covSet2 <- createCovariateSettings(useDemographicsGender = T, 
                                   useDemographicsAgeGroup = T, 
                                   useConditionGroupEraAnyTimePrior = T,
                                   useDrugGroupEraAnyTimePrior = F)
covariateSettingList <- list(covSet1, covSet2)
```                                   
# Model Settings                                    
The model settings requires running the setModel functions for the machine learning models of interest and specifying the hyper-parameter search and then combining these into a list.  For example, if we wanted to try a logistic regression, gradient boosting machine and ada boost model then:

```{r tidy=FALSE,eval=FALSE}
gbm <- setGradientBoostingMachine()
lr <- setLassoLogisticRegression()
ada <- setAdaBoost()

modelList <- list(gbm, lr, ada)
```   

# Creating the model analysis list   
To create the complete plp model settings use `createPlpModelSettings` to combine the population, covariate and model settings.

```{r tidy=FALSE,eval=FALSE}
modelAnalysisList <- createPlpModelSettings(modelList = modelList, 
                                   covariateSettingList = covariateSettingList,
                                   populationSettingList = populationSettingList)
```  

# Running the multiple prediction patient-level-prediction
As we will be downloading loads of data in the multiple plp analysis it is useful to set fftempdir to a directory with write access and plenty of space.
`options(fftempdir = 'T:/fftemp')`

To run the study requires setting up a connectionDetails object

```{r tidy=FALSE,eval=FALSE}
dbms <- "your dbms"
user <- "your username"
pw <- "your password"
server <- "your server"
port <- "your port"

connectionDetails <- DatabaseConnector::createConnectionDetails(dbms = dbms,
                                                                server = server,
                                                                user = user,
                                                                password = pw,
                                                                port = port)
```

Next you need to specify the cdmDatabaseSchema where your cdm database is found and workDatabaseSchema where your target population and outcome cohorts are.

```{r tidy=FALSE,eval=FALSE}
cdmDatabaseSchema <- "your cdmDatabaseSchema"
workDatabaseSchema <- "your workDatabaseSchema"
```

Now you can run the multiple patient-level prediction analysis by specifying the target cohort ids and outcome ids 

```{r tidy=FALSE,eval=FALSE}
allresults <- runPlpAnalyses(connectionDetails = connectionDetails,
                           cdmDatabaseSchema = cdmDatabaseSchema,
                           oracleTempSchema = cdmDatabaseSchema,
                           cohortDatabaseSchema = workDatabaseSchema,
                           cohortTable = "your cohort table",
                           outcomeDatabaseSchema = workDatabaseSchema,
                           outcomeTable = "your cohort table",
                           cdmVersion = 5,
                           outputFolder = "./PlpMultiOutput",
                           modelAnalysisList = modelAnalysisList,
                           cohortIds = c(2484,6970),
                           cohortNames = c('visit 2010','test cohort'),
                           outcomeIds = c(7331,5287),
                           outcomeNames =  c('outcome 1','outcome 2'),
                           maxSampleSize = NULL,
                           minCovariateFraction = 0,
                           normalizeData = T,
                           testSplit = "person",
                           testFraction = 0.25,
                           splitSeed = NULL,
                           nfold = 3,
                           verbosity = "INFO")
```
This will then save all the plpData objects from the study into "./PlpMultiOutput/plpData", the populations for the analysis into "./PlpMultiOutput/population" and the results into "./PlpMultiOutput/Result".  The csv named settings.csv found in "./PlpMultiOutput" has a row for each prediction model developed and points to the plpData and popualtion used for the model development, it also has descriptions of the cohorts and settings if these are input by the user.

# Validating the multiple prediction patient-level-prediction results
To validate all the models on new data and cohorts run (where analysesLocation is the output location you input for runPlpAnalyses() and the suggested outputLocation is the subdirectory 'validation' in this location):
```{r tidy=FALSE,eval=FALSE}
val <- evaluateMultiplePlp(analysesLocation = "./PlpMultiOutput",
                                outputLocation = "./PlpMultiOutput/validation",
                                connectionDetails = connectionDetails, 
                                validationSchemaTarget = list('new_database_1.dbo',
                                                              'new_database_2.dbo'),
                                validationSchemaOutcome = list('new_database_1.dbo',
                                                              'new_database_2.dbo'),
                                validationSchemaCdm = list('new_database_1.dbo',
                                                              'new_database_2.dbo'), 
                                databaseNames = c('database1','database2'),
                                validationTableTarget = 'your new cohort table',
                                validationTableOutcome = 'your new cohort table')
```
This then saves the external validation results in the validation folder of the main study (the outputLocation you used in runPlpAnalyses).

# Viewing the multiple prediction patient-level-prediction results
To view the results for the multiple prediction analysis:
```{r tidy=FALSE,eval=FALSE}
viewMultiplePlp(analysesLocation="./PlpMultiOutput")
```
If the validation directory in "./PlpMultiOutput" has results, the external validation will also be displayed.