---
title: "Constrained predictors"
author: "Jenna Reps"
date: '`r Sys.Date()`'
header-includes:
    - \usepackage{fancyhdr}
    - \pagestyle{fancy}
    - \fancyhead{}
    - \fancyhead[CO,CE]{Installation Guide}
    - \fancyfoot[CO,CE]{PatientLevelPrediction Package Version `r    utils::packageVersion("PatientLevelPrediction")`}
    - \fancyfoot[LE,RO]{\thepage}
    - \renewcommand{\headrulewidth}{0.4pt}
    - \renewcommand{\footrulewidth}{0.4pt}
output:
  pdf_document:
    includes:
      in_header: preamble.tex
    number_sections: yes
    toc: yes
  word_document:
    toc: yes
  html_document:
    number_sections: yes
    toc: yes
---
<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Constrainted Predictors}
-->

## Constrained Predictors


### How to use the PhenotypeLibrary R package
Here we provide a set of phenotypes that can be used as predictors in prediction models or best practice research.

These phenotypes can be extracted from the PhenotypeLibrary R package.  To install the R package run:

```{r echo = T}
remotes::install_github('ohdsi/PhenotypeLibrary')
```


To extract the cohort definition for Alcoholism with an id of 1165, just run:

```{r echo = T}
PhenotypeLibrary::getPlCohortDefinitionSet(1165)
```

in general you can extract all the cohorts by running:

```{r echo = T}
phenotypeDefinitions <- PhenotypeLibrary::getPlCohortDefinitionSet(1152:1215)
```

### The full set of predictor phenotypes

<table>

<tr>
<th>
Phenotype Name
</th>
<th>
Disorder classification
</th>
<th>
OHDSI Phenotype library ID
</th>
</tr>

<tr>
<td>
Alcoholism
</td>
<td>
Behavioral 
</td>
<td>
1165
</td>
</tr>

<tr>
<td>
Smoking
</td>
<td>
Behavioral 
</td>
<td>
1166
</td>
</tr>

<tr>
<td>
Anemia
</td>
<td>
Blood
</td>
<td>
1188
</td>
</tr>

<tr>
<td>
Osteoarthritis
</td>
<td>
Bone
</td>
<td>
1184
</td>
</tr>

<tr>
<td>
Osteoporosis
</td>
<td>
Bone
</td>
<td>
1185
</td>
</tr>

<tr>
<td>
Cancer
</td>
<td>
Cancer
</td>
<td>
1215
</td>
</tr>

<tr>
<td>
Atrial fibrillation
</td>
<td>
Cardiovascular
</td>
<td>
1160
</td>
</tr>

<tr>
<td>
Congestive heart failure
</td>
<td>
Cardiovascular
</td>
<td>
1154
</td>
</tr>


<tr>
<td>
Coronary artery disease
</td>
<td>
Cardiovascular
</td>
<td>
1162
</td>
</tr>

<tr>
<td>
Heart valve disorder
</td>
<td>
Cardiovascular
</td>
<td>
1172
</td>
</tr>


<tr>
<td>
Hyperlipidemia
</td>
<td>
Cardiovascular
</td>
<td>
1170
</td>
</tr>


<tr>
<td>
Hypertension
</td>
<td>
Cardiovascular
</td>
<td>
1198
</td>
</tr>


<tr>
<td>
Angina
</td>
<td>
Cardiovascular
</td>
<td>
1159
</td>
</tr>


<tr>
<td>
Skin Ulcer
</td>
<td>
Debility
</td>
<td>
1168
</td>
</tr>

<tr>
<td>
Diabetes type 1
</td>
<td>
Endocrine
</td>
<td>
1193
</td>
</tr>

<tr>
<td>
Diabetes type 2
</td>
<td>
Endocrine
</td>
<td>
1194
</td>
</tr>


<tr>
<td>
Hypothyroidism
</td>
<td>
Endocrine
</td>
<td>
1171
</td>
</tr>


<tr>
<td>
Obesity
</td>
<td>
Endocrine
</td>
<td>
1179
</td>
</tr>


<tr>
<td>
Gastroesophageal reflux disease (GERD)
</td>
<td>
GI
</td>
<td>
1178
</td>
</tr>


<tr>
<td>
Gastrointestinal (GI) bleed 
</td>
<td>
GI
</td>
<td>
1197
</td>
</tr>


<tr>
<td>
Inflammatory bowel disorder (IBD) 
</td>
<td>
GI/Rheumatology
</td>
<td>
1180
</td>
</tr>


<tr>
<td>
Hormonal contraceptives
</td>
<td>
Gynecologic 
</td>
<td>
1190
</td>
</tr>


<tr>
<td>
Antibiotics Aminoglycosides
</td>
<td>
Infection
</td>
<td>
1201
</td>
</tr>

<tr>
<td>
Antibiotics Carbapenems
</td>
<td>
Infection
</td>
<td>
1202
</td>
</tr>

<tr>
<td>
Antibiotics Cephalosporins
</td>
<td>
Infection
</td>
<td>
1203
</td>
</tr>


<tr>
<td>
Antibiotics Fluoroquinolones
</td>
<td>
Infection
</td>
<td>
1204
</td>
</tr>

<tr>
<td>
Antibiotics Glycopeptides and lipoglycopeptides
</td>
<td>
Infection
</td>
<td>
1205
</td>
</tr>

<tr>
<td>
Antibiotics Macrolides
</td>
<td>
Infection
</td>
<td>
1206
</td>
</tr>

<tr>
<td>
Antibiotics Monobactams
</td>
<td>
Infection
</td>
<td>
1207
</td>
</tr>


<tr>
<td>
Antibiotics Oxazolidinones
</td>
<td>
Infection
</td>
<td>
1208
</td>
</tr>

<tr>
<td>
Antibiotics Penicillins
</td>
<td>
Infection
</td>
<td>
1209
</td>
</tr>


<tr>
<td>
Antibiotics Polypeptides
</td>
<td>
Infection
</td>
<td>
1210
</td>
</tr>

<tr>
<td>
Antibiotics Rifamycins
</td>
<td>
Infection
</td>
<td>
1211
</td>
</tr>

<tr>
<td>
Antibiotics Sulfonamides
</td>
<td>
Infection
</td>
<td>
1212
</td>
</tr>

<tr>
<td>
Antibiotics Streptogramins
</td>
<td>
Infection
</td>
<td>
1213
</td>
</tr>

<tr>
<td>
Antibiotics Tetracyclines
</td>
<td>
Infection
</td>
<td>
1214
</td>
</tr>

<tr>
<td>
Pneumonia
</td>
<td>
Infection/Respiratory 
</td>
<td>
1199
</td>
</tr>

<tr>
<td>
Sepsis
</td>
<td>
Infection
</td>
<td>
1176
</td>
</tr>

<tr>
<td>
Urinary tract infection (UTI)
</td>
<td>
Infection
</td>
<td>
1186
</td>
</tr>


<tr>
<td>
Hepatitis
</td>
<td>
Liver
</td>
<td>
1169
</td>
</tr>


<tr>
<td>
Anxiety
</td>
<td>
Mood
</td>
<td>
1189
</td>
</tr>


<tr>
<td>
Depression (MDD)
</td>
<td>
Mood
</td>
<td>
1161
</td>
</tr>

<tr>
<td>
Psychotic disorder
</td>
<td>
Mood
</td>
<td>
1175
</td>
</tr>


<tr>
<td>
Antiepileptics (pain)
</td>
<td>
Neurology/Pain
</td>
<td>
1183
</td>
</tr>

<tr>
<td>
Seizure
</td>
<td>
Neurology
</td>
<td>
1153
</td>
</tr>


<tr>
<td>
Hemorrhagic stroke
</td>
<td>
Neurology/Vascular
</td>
<td>
1156
</td>
</tr>


<tr>
<td>
Non-hemorrhagic stroke
</td>
<td>
Neurology/Vascular
</td>
<td>
1155
</td>
</tr>

<tr>
<td>
Acetaminophen prescription
</td>
<td>
Pain/Infection
</td>
<td>
1187
</td>
</tr>



<tr>
<td>
Low back pain
</td>
<td>
Pain
</td>
<td>
1173
</td>
</tr>


<tr>
<td>
Neuropathy
</td>
<td>
Pain/Neurology
</td>
<td>
1174
</td>
</tr>


<tr>
<td>
Opioids
</td>
<td>
Pain
</td>
<td>
1182
</td>
</tr>

<tr>
<td>
Acute kidney injury
</td>
<td>
Kidney
</td>
<td>
1163
</td>
</tr>


<tr>
<td>
Chronic kidney disease
</td>
<td>
Kidney
</td>
<td>
1191
</td>
</tr>

<tr>
<td>
Asthma
</td>
<td>
Respiratory
</td>
<td>
1164
</td>
</tr>

<tr>
<td>
Chronic obstructive pulmonary disorder (COPD)
</td>
<td>
Respiratory
</td>
<td>
1192
</td>
</tr>

<tr>
<td>
Dyspnea
</td>
<td>
Respiratory
</td>
<td>
1195
</td>
</tr>

<tr>
<td>
Respiratory failure
</td>
<td>
Respiratory
</td>
<td>
1177
</td>
</tr>


<tr>
<td>
Sleep apnea
</td>
<td>
Respiratory
</td>
<td>
1167
</td>
</tr>

<tr>
<td>
Rheumatoid arthritis
</td>
<td>
Rheumatology
</td>
<td>
1200
</td>
</tr>


<tr>
<td>
Steroids
</td>
<td>
Rheumatology/Pain/Pulmonary
</td>
<td>
1181
</td>
</tr>


<tr>
<td>
Peripheral vascular disease
</td>
<td>
Vascular
</td>
<td>
1157
</td>
</tr>


<tr>
<td>
Aspirin
</td>
<td>
Vascular
</td>
<td>
1158
</td>
</tr>

<tr>
<td>
Deep vein thrombosis (DVT)
</td>
<td>
Vascular
</td>
<td>
1152
</td>
</tr>

<tr>
<td>
Edema
</td>
<td>
Vascular
</td>
<td>
1196
</td>
</tr>

<tr>
<td>
Inpatient visit
</td>
<td>
NA
</td>
<td>
NA
</td>
</tr>

</table>
    