---
title: "Constrained Predictors"
author: "Jenna Reps"
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Constrained Predictors}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

## Constrained Predictors

### How to use the PhenotypeLibrary R package

Here we provide a set of phenotypes that can be used as predictors in prediction models or best practice research.

These phenotypes can be extracted from the PhenotypeLibrary R package. To install the R package run:

```{r, echo = TRUE, message = FALSE, warning = FALSE, tidy = FALSE, eval=FALSE}
remotes::install_github("OHDSI/PhenotypeLibrary")
```

To extract the cohort definition for Alcoholism with an id of 1165, just run:

```{r echo = TRUE, message = FALSE, warning = FALSE, tidy = FALSE, eval = FALSE}
PhenotypeLibrary::getPlCohortDefinitionSet(1165)
```

in general you can extract all the cohorts by running:

```{r echo = TRUE, message = FALSE, warning = FALSE, tidy = FALSE, eval = FALSE}
phenotypeDefinitions <- PhenotypeLibrary::getPlCohortDefinitionSet(1152:1215)
```

### The full set of predictor phenotypes

```{r show-table, eval=TRUE, echo=FALSE, tidy = FALSE}
df <- tibble::tribble(
  ~"Phenotype Name", ~"Disorder classification", ~"OHDSI Phenotype library ID",
  "Alcoholism", "Behavioral", 1165,
  "Smoking", "Behavioral", 1166,
  "Anemia", "Blood", 1188,
  "Osteoarthritis", "Bone", 1184,
  "Osteoporosis", "Bone", 1185,
  "Cancer", "Cancer", 1215,
  "Atrial fibrillation", "Cardiovascular", 1160,
  "Congestive heart failure", "Cardiovascular", 1154,
  "Coronary artery disease", "Cardiovascular", 1162,
  "Heart valve disorder", "Cardiovascular", 1172,
  "Hyperlipidemia", "Cardiovascular", 1170,
  "Hypertension", "Cardiovascular", 1198,
  "Angina", "Cardiovascular", 1159,
  "Skin Ulcer", "Debility", 1168,
  "Diabetes type 1", "Endocrine", 1193,
  "Diabetes type 2", "Endocrine", 1194,
  "Hypothyroidism", "Endocrine", 1171,
  "Obesity", "Endocrine", 1179,
  "Gastroesophageal reflux disease (GERD)", "GI", 1178,
  "Gastrointestinal (GI) bleed", "GI", 1197,
  "Inflammatory bowel disorder (IBD)", "GI/Rheumatology", 1180,
  "Hormonal contraceptives", "Gynecologic", 1190,
  "Antibiotics Aminoglycosides", "Infection", 1201,
  "Antibiotics Carbapenems", "Infection", 1202,
  "Antibiotics Cephalosporins", "Infection", 1203,
  "Antibiotics Fluoroquinolones", "Infection", 1204,
  "Antibiotics Glycopeptides and lipoglycopeptides", "Infection", 1205,
  "Antibiotics Macrolides", "Infection", 1206,
  "Antibiotics Monobactams", "Infection", 1207,
  "Antibiotics Oxazolidinones", "Infection", 1208,
  "Antibiotics Penicillins", "Infection", 1209,
  "Antibiotics Polypeptides", "Infection", 1210,
  "Antibiotics Rifamycins", "Infection", 1211,
  "Antibiotics Sulfonamides", "Infection", 1212,
  "Antibiotics Streptogramins", "Infection", 1213,
  "Antibiotics Tetracyclines", "Infection", 1214,
  "Pneumonia", "Infection/Respiratory", 1199,
  "Sepsis", "Infection", 1176,
  "Urinary tract infection (UTI)", "Infection", 1186,
  "Hepatitis", "Liver", 1169,
  "Anxiety", "Mood", 1189,
  "Depression (MDD)", "Mood", 1161,
  "Psychotic disorder", "Mood", 1175,
  "Antiepileptics (pain)", "Neurology/Pain", 1183,
  "Seizure", "Neurology", 1153,
  "Hemorrhagic stroke", "Neurology/Vascular", 1156,
  "Non-hemorrhagic stroke", "Neurology/Vascular", 1155,
  "Acetaminophen prescription", "Pain/Infection", 1187,
  "Low back pain", "Pain", 1173,
  "Neuropathy", "Pain/Neurology", 1174,
  "Opioids", "Pain", 1182,
  "Acute kidney injury", "Kidney", 1163,
  "Chronic kidney disease", "Kidney", 1191,
  "Asthma", "Respiratory", 1164,
  "Chronic obstructive pulmonary disorder (COPD)", "Respiratory", 1192,
  "Dyspnea", "Respiratory", 1195,
  "Respiratory failure", "Respiratory", 1177,
  "Sleep apnea", "Respiratory", 1167,
  "Rheumatoid arthritis", "Rheumatology", 1200,
  "Steroids", "Rheumatology/Pain/Pulmonary", 1181,
  "Peripheral vascular disease", "Vascular", 1157,
  "Aspirin", "Vascular", 1158,
  "Deep vein thrombosis (DVT)", "Vascular", 1152,
  "Edema", "Vascular", 1196
)

#Build a reactableTheme that *reads* Bootstrap CSS vars
bs_table_theme <- reactable::reactableTheme(
  backgroundColor = "var(--bs-table-bg)",
  color           = "var(--bs-table-color)",
  searchInputStyle = list(
    backgroundColor = "var(--bs-table-bg)",
    color           = "var(--bs-table-color)"
  )
)
reactable::reactable(
  df,
  theme        = bs_table_theme,
  searchable   = TRUE,        # add the search box
  defaultPageSize = 30,       # rows per page
  highlight     = TRUE,       # highlight row under cursor
  bordered      = TRUE,       # add table borders
  striped       = TRUE,       # zebra-striped rows
  columns = list(
    `Phenotype Name` = reactable::colDef(
      name     = "Phenotype",
      minWidth = 200
    ),
    `Disorder classification` = reactable::colDef(
      name     = "Category"
    ),
    `OHDSI Phenotype library ID` = reactable::colDef(
      name            = "Library ID",
      defaultSortOrder = "desc",
      align           = "right"
    )
  )
)
```
