---
title: "Constrained predictors"
author: "Jenna Reps"
date: '`r Sys.Date()`'
header-includes:
    - \usepackage{fancyhdr}
    - \pagestyle{fancy}
    - \fancyhead{}
    - \fancyhead[CO,CE]{Installation Guide}
    - \fancyfoot[CO,CE]{PatientLevelPrediction Package Version `r    utils::packageVersion("PatientLevelPrediction")`}
    - \fancyfoot[LE,RO]{\thepage}
    - \renewcommand{\headrulewidth}{0.4pt}
    - \renewcommand{\footrulewidth}{0.4pt}
output:
  pdf_document:
    includes:
      in_header: preamble.tex
    number_sections: yes
    toc: yes
  word_document:
    toc: yes
  html_document:
    number_sections: yes
    toc: yes
---

```{=html}
<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Constrainted Predictors}
-->
```
## Constrained Predictors

### How to use the PhenotypeLibrary R package

Here we provide a set of phenotypes that can be used as predictors in prediction models or best practice research.

These phenotypes can be extracted from the PhenotypeLibrary R package. To install the R package run:

```{r, echo = TRUE, message = FALSE, warning = FALSE, tidy = FALSE, eval=FALSE}
remotes::install_github('ohdsi/PhenotypeLibrary')
```

To extract the cohort definition for Alcoholism with an id of 1165, just run:

```{r echo = TRUE, message = FALSE, warning = FALSE, tidy = FALSE, eval = FALSE}
PhenotypeLibrary::getPlCohortDefinitionSet(1165)
```

in general you can extract all the cohorts by running:

```{r echo = TRUE, message = FALSE, warning = FALSE, tidy = FALSE, eval = FALSE}
phenotypeDefinitions <- PhenotypeLibrary::getPlCohortDefinitionSet(1152:1215)
```

### The full set of predictor phenotypes

| Phenotype Name                                  | Disorder classification     | OHDSI Phenotype library ID |
|--------------------------------|--------------------|-------------------|
| Alcoholism                                      | Behavioral                  | 1165                       |
| Smoking                                         | Behavioral                  | 1166                       |
| Anemia                                          | Blood                       | 1188                       |
| Osteoarthritis                                  | Bone                        | 1184                       |
| Osteoporosis                                    | Bone                        | 1185                       |
| Cancer                                          | Cancer                      | 1215                       |
| Atrial fibrillation                             | Cardiovascular              | 1160                       |
| Congestive heart failure                        | Cardiovascular              | 1154                       |
| Coronary artery disease                         | Cardiovascular              | 1162                       |
| Heart valve disorder                            | Cardiovascular              | 1172                       |
| Hyperlipidemia                                  | Cardiovascular              | 1170                       |
| Hypertension                                    | Cardiovascular              | 1198                       |
| Angina                                          | Cardiovascular              | 1159                       |
| Skin Ulcer                                      | Debility                    | 1168                       |
| Diabetes type 1                                 | Endocrine                   | 1193                       |
| Diabetes type 2                                 | Endocrine                   | 1194                       |
| Hypothyroidism                                  | Endocrine                   | 1171                       |
| Obesity                                         | Endocrine                   | 1179                       |
| Gastroesophageal reflux disease (GERD)          | GI                          | 1178                       |
| Gastrointestinal (GI) bleed                     | GI                          | 1197                       |
| Inflammatory bowel disorder (IBD)               | GI/Rheumatology             | 1180                       |
| Hormonal contraceptives                         | Gynecologic                 | 1190                       |
| Antibiotics Aminoglycosides                     | Infection                   | 1201                       |
| Antibiotics Carbapenems                         | Infection                   | 1202                       |
| Antibiotics Cephalosporins                      | Infection                   | 1203                       |
| Antibiotics Fluoroquinolones                    | Infection                   | 1204                       |
| Antibiotics Glycopeptides and lipoglycopeptides | Infection                   | 1205                       |
| Antibiotics Macrolides                          | Infection                   | 1206                       |
| Antibiotics Monobactams                         | Infection                   | 1207                       |
| Antibiotics Oxazolidinones                      | Infection                   | 1208                       |
| Antibiotics Penicillins                         | Infection                   | 1209                       |
| Antibiotics Polypeptides                        | Infection                   | 1210                       |
| Antibiotics Rifamycins                          | Infection                   | 1211                       |
| Antibiotics Sulfonamides                        | Infection                   | 1212                       |
| Antibiotics Streptogramins                      | Infection                   | 1213                       |
| Antibiotics Tetracyclines                       | Infection                   | 1214                       |
| Pneumonia                                       | Infection/Respiratory       | 1199                       |
| Sepsis                                          | Infection                   | 1176                       |
| Urinary tract infection (UTI)                   | Infection                   | 1186                       |
| Hepatitis                                       | Liver                       | 1169                       |
| Anxiety                                         | Mood                        | 1189                       |
| Depression (MDD)                                | Mood                        | 1161                       |
| Psychotic disorder                              | Mood                        | 1175                       |
| Antiepileptics (pain)                           | Neurology/Pain              | 1183                       |
| Seizure                                         | Neurology                   | 1153                       |
| Hemorrhagic stroke                              | Neurology/Vascular          | 1156                       |
| Non-hemorrhagic stroke                          | Neurology/Vascular          | 1155                       |
| Acetaminophen prescription                      | Pain/Infection              | 1187                       |
| Low back pain                                   | Pain                        | 1173                       |
| Neuropathy                                      | Pain/Neurology              | 1174                       |
| Opioids                                         | Pain                        | 1182                       |
| Acute kidney injury                             | Kidney                      | 1163                       |
| Chronic kidney disease                          | Kidney                      | 1191                       |
| Asthma                                          | Respiratory                 | 1164                       |
| Chronic obstructive pulmonary disorder (COPD)   | Respiratory                 | 1192                       |
| Dyspnea                                         | Respiratory                 | 1195                       |
| Respiratory failure                             | Respiratory                 | 1177                       |
| Sleep apnea                                     | Respiratory                 | 1167                       |
| Rheumatoid arthritis                            | Rheumatology                | 1200                       |
| Steroids                                        | Rheumatology/Pain/Pulmonary | 1181                       |
| Peripheral vascular disease                     | Vascular                    | 1157                       |
| Aspirin                                         | Vascular                    | 1158                       |
| Deep vein thrombosis (DVT)                      | Vascular                    | 1152                       |
| Edema                                           | Vascular                    | 1196                       |
| Inpatient visit                                 | NA                          | NA                         |
