---
title: "Patient-Level Prediction V5.1.0 Installation Guide"
author: "Jenna Reps, Peter R. Rijnbeek"
date: '`r Sys.Date()`'
header-includes:
    - \usepackage{fancyhdr}
    - \pagestyle{fancy}
    - \fancyhead{}
    - \fancyhead[CO,CE]{Installation Guide}
    - \fancyfoot[CO,CE]{PatientLevelPrediction Package Version `r    utils::packageVersion("PatientLevelPrediction")`}
    - \fancyfoot[LE,RO]{\thepage}
    - \renewcommand{\headrulewidth}{0.4pt}
    - \renewcommand{\footrulewidth}{0.4pt}
output:
  pdf_document:
    includes:
      in_header: preamble.tex
    number_sections: yes
    toc: yes
  word_document:
    toc: yes
  html_document:
    number_sections: yes
    toc: yes
---
<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Installation guide}
-->

# Introduction
This vignette describes how you need to install the Observational Health Data Sciencs and Informatics (OHDSI) [`PatientLevelPrediction`](http://github.com/OHDSI/PatientLevelPrediction) package under Windows, Mac, and Linux.

# Software Prerequisites

## Windows Users
Under Windows the OHDSI Patient Level Prediction (PLP) package requires installing:

* R (https://cran.cnr.berkeley.edu/ ) \- (R >= 3.3.0, but latest is recommended)
* Rstudio (https://www.rstudio.com/  )
* Java (http://www.java.com )
* RTools (https://cran.r-project.org/bin/windows/Rtools/) 

## Mac/Linux Users
Under Mac and Linux the OHDSI Patient Level Prediction (PLP) package requires installing:

* R (https://cran.cnr.berkeley.edu/ ) \- (R >= 3.3.0, but latest is recommended)
* Rstudio (https://www.rstudio.com/  )
* Java (http://www.java.com )
* Xcode command line tools(run in terminal: xcode-select --install) [MAC USERS ONLY]


# Installing the Package
The preferred way to install the package is by using `remotes`, which will automatically install the latest release and all the latest dependencies.

If you do not want the official release you could install the bleading edge version of the package (latest develop branch).

Note that the latest develop branch could contain bugs, please report them to us if you experience problems.

## Installing PatientLevelPrediction using remotes
To install using `remotes` run:
```{r, echo = TRUE, message = FALSE, warning = FALSE,tidy=FALSE,eval=FALSE}
install.packages("remotes")
remotes::install_github("OHDSI/FeatureExtraction")
remotes::install_github("OHDSI/PatientLevelPrediction")
```

When installing make sure to close any other Rstudio sessions that are using `PatientLevelPrediction` or any dependency.  Keeping Rstudio sessions open can cause locks that prevent the package installing.

# Creating Python Reticulate Environment

Many of the classifiers in the `PatientLevelPrediction` use a Python back end.  To set up a python environment run:

```{r, echo = TRUE, message = FALSE, warning = FALSE,tidy=FALSE,eval=FALSE}
library(PatientLevelPrediction)
reticulate::install_miniconda()
configurePython(envname='r-reticulate', envtype='conda')

```

Some of the less frequently used classifiers are not installed during this set-up to add them run:

For GBM survival:
```{r, echo = TRUE, message = FALSE, warning = FALSE,tidy=FALSE,eval=FALSE}
reticulate::conda_install(envname='r-reticulate', packages = c('scikit-survival'), forge = TRUE, pip = FALSE, pip_ignore_installed = TRUE, conda = "auto", channel = 'sebp')

```

# Testing installation
To test whether the package is installed correctly, using the test script in '/extras', run:
```{r, echo = TRUE, message = FALSE, warning = FALSE,tidy=FALSE,eval=FALSE}

# load the checkPlpInstallation function
library(devtools)
source_url('https://raw.github.com/OHDSI/PatientLevelPrediction/issue242/extras/checkPlpInstallation.R')

# set up the database connection details
library(DatabaseConnector)
connectionDetails <- createConnectionDetails(
  dbms = 'sql_server', 
  user = 'username', 
  password = 'hidden', 
  server = 'your server', 
  port = 'your port'
  )

# run the test
checkPlpInstallation(
  connectionDetails = connectionDetails, 
  python = T
  )
```

To test the installation (excluding python) run:
```{r, echo = TRUE, message = FALSE, warning = FALSE,tidy=FALSE,eval=FALSE}

checkPlpInstallation(
  connectionDetails = connectionDetails, 
  python = F
  )
```

The check can take a while to run since it will build the following models in sequence on simulated data: Logistic Regression, RandomForest, MLP, AdaBoost, Decision Tree, Naive Bayes, KNN, Gradient Boosting.
Moreover, it will test the database connection. 

# Installation issues
Installation issues need to be posted in our issue tracker:
http://github.com/OHDSI/PatientLevelPrediction/issues

The list below provides solutions for some common issues:

1. If you have an error when trying to install a package in R saying **'Dependancy X not available ...'** then this can sometimes be fixed by running `install.packages('X')` and then once that completes trying to reinstall the package that had the error.

2. I have found that using the github `remotes`` to install packages can be impacted if you have **multiple R sessions** open as one session with a library open can cause the library to be locked and this can prevent an install of a package that depends on that library.


## Common issues
### python environment Mac/linux users: 
to make sure R uses the r-reticulate python environment you may need to edit your .Rprofile with the location of the python binary for the PLP environment.    Edit the .Rprofile by running: 
```{r eval=FALSE}
usethis::edit_r_profile()
``` 
and add  
```{r eval=FALSE} 
Sys.setenv(PATH = paste("your python bin location", Sys.getenv("PATH"), sep=":"))
``` 
to the file then save.  Where your python bin location is the location returned by 
```{r eval=FALSE} 
reticulate::conda_list() 
``` 
e.g., My PLP virtual environment location was /anaconda3/envs/PLP/bin/python so I added:  
Sys.setenv(PATH = paste("/anaconda3/envs/PLP/bin", Sys.getenv("PATH"), sep=":"))

# Acknowledgments

Considerable work has been dedicated to provide the `PatientLevelPrediction` package.

```{r tidy=TRUE,eval=TRUE}
citation("PatientLevelPrediction")
```

**Please reference this paper if you use the PLP Package in your work:**

[Reps JM, Schuemie MJ, Suchard MA, Ryan PB, Rijnbeek PR. Design and implementation of a standardized framework to generate and evaluate patient-level prediction models using observational healthcare data. J Am Med Inform Assoc. 2018;25(8):969-975.](http://dx.doi.org/10.1093/jamia/ocy032)


This work is supported in part through the National Science Foundation grant IIS 1251151.